**Measuring Depression and Anxiety with 4 items? Adaptation of the PHQ-4 to increase its Sensitivity to Subclinical Variability**

### 2025-06-06 paper/5_Collabra/manuscript.qmd
replace STAIT with STAI (couldn't be bothered to change the figures)

This brief report aims at testing the possibility of enhancing - with minimal changes to the original scale - the PHQ-4 sensitivity to mild mood level inflections. In the first study, we will evaluate whether the new response option is prevalently used by participants, and whether it does capture a specific part of the construct. In the second study, we will compare the refined PHQ-4 version to the original one in terms of sensitivity to subclinical variability in depression, using the Beck Depression Inventory [BDI-II, @beck1996beck] and the State Trait Anxiety Inventory [STAI-5, @zsido2020development] as our ground-truth measures of depression and anxiety.

@@

For the refined version, we added a "Once or twice" option between "Not at all" and "Several days"in order to better capture potential mild mood inflections [see @dobson1979equidistant for the choice of the label]. This new option was scored as 0.5 to preserve the same scoring as the original version.

@@

The reliability of the anxiety ($Cronbach's~\alpha = 0.903$; RMSEA = 0.031) and depression ($Cronbach's~\alpha = 0.841$; RMSEA = 0.044) subscales is excellent. The proportion of response types stratified by item (see @fig-one) shows that the new "Once or twice" option was the most prevalent response for all items (on average selected in 29.12\% of cases).

@@

Item Response Theory (IRT) provides insights into how well items and responses capture an underlying latent trait $\theta$. For each of the subscales, we fitted a unidimensional graded response model [GRM, @samejima1997graded]. For anxiety, the two items captured 89.2\% of the variance of the latent anxiety dimension ($\theta_{anxiety}$). The discrimination parameters suggested that the first item was less precise ($\alpha = 3.42$) than the second item ($\alpha = 12.55$) in its ability to discriminate between various levels of anxiety (i.e., each response on the second item covers a more exclusive range of $\theta_{anxiety}$, as can be seen in @fig-one). The two depression items captured 82.8\% of the variance of its latent trait ($\theta_{depression}$), and the opposite pattern was found: the first item had a higher precision ($\alpha = 16.46$) than the second ($\alpha = 2.41$). However, it is important to note that the "less precise" items were also the ones covering a larger portion of the latent space (being more sensitive especially on the lower end of the spectrum), offering an interesting trade-off between sensitivity and precision. Most importantly for our objective, the added "Once or twice" option did cover a selective and unique portion of the latent space.

@@

Participants were randomly assigned to complete either the original or refined version of the PHQ-4, which included one additional response option ("Once or twice") scored 0.5 (creating more possible total scores - 0.5, 1.5, 2.5, etc.).

@@

The short version of the State-Trait Anxiety Inventory [STAI-5, @zsido2020development] was used as a ground truth measure of anxiety. This abridged version of the STAI [@spielberger1970manual] includes 5 items rated on a 4-point Likert scale. Changes were made in the instructions from asking *"how participants feel right now"* to *"over the past 2 weeks"* to keep it consistent with the instructions of the PHQ-4 and BDI-II. A general score of anxiety was computed by averaging all the items.

Participants were also asked to complete two questionnaires of interoception, namely the Interoceptive Accuracy Scale [IAS - 21 items rated on analog scales, @murphy2020testing] and the Multidimensional Assessment of Interoceptive Awareness [MAIA-2 - 37 items, @mehling2018multidimensional].

@@

Finally, although we used a stricter criterion for classifying participants as having a depression or an anxiety disorder by restricting it to participants also reporting undergoing a medical treatment, it was still based on self-reported data. Studies in controlled clinical settings are needed to confirm the potential benefits of the refined PHQ-4 in mood disorders detection accuracy.

@@

>>>>

This brief report aims at testing the possibility of enhancing - with minimal changes to the original scale - the PHQ-4 sensitivity to mild mood level inflections. In the first study, we will evaluate whether the new response option is prevalently used by participants, and whether it does capture a specific part of the construct. In the second study, we will compare the refined PHQ-4 version to the original one in terms of sensitivity to subclinical variability in depression, using the Beck Depression Inventory [BDI-II, @beck1996beck] and the Trait scale of the State-Trait Anxiety Inventory [STAI-5, @zsido2020development] as our ground-truth measures of depression and anxiety.

@@

For the refined version, we added a "Once or twice" option between "Not at all" and "Several days"in order to better capture potential mild mood inflections [see @dobson1979equidistant for the choice of the label].

@@

The consistency of the anxiety ($Cronbach's~\alpha = 0.903$) and depression ($Cronbach's~\alpha = 0.841$) subscales is excellent (responses were treated as ordinal). The proportion of response types stratified by item (see @fig-one) shows that the new "Once or twice" option was the most prevalent response for all items (on average selected in 29.12\% of cases).

@@

Item Response Theory (IRT) provides insights into how well items and responses capture an underlying latent trait $\theta$. For each of the subscales, we fitted a unidimensional graded response model [GRM, @samejima1997graded]. For anxiety, the two items captured 89.2\% of the variance (RMSEA = 0.031) of the latent anxiety dimension ($\theta_{anxiety}$). The discrimination parameters suggested that the first item was less precise ($\alpha = 3.42$) than the second item ($\alpha = 12.55$) in its ability to discriminate between various levels of anxiety (i.e., each response on the second item covers a more exclusive range of $\theta_{anxiety}$, as can be seen in @fig-one). The two depression items captured 82.8\% of the variance (RMSEA = 0.044) of its latent trait ($\theta_{depression}$), and the opposite pattern was found: the first item had a higher precision ($\alpha = 16.46$) than the second ($\alpha = 2.41$). However, it is important to note that the "less precise" items were also the ones covering a larger portion of the latent space (being more sensitive especially on the lower end of the spectrum), offering an interesting trade-off between sensitivity and precision. Most importantly for our objective, the added "Once or twice" option did cover a selective and unique portion of the latent space.

@@

Participants were randomly assigned to complete either the original or refined version of the PHQ-4, which included one additional response option ("Once or twice"), scored 0.5 to preserve the same scoring as the original version (creating more possible total scores - 0.5, 1.5, 2.5, etc.).

@@

The short version of the trait subscale of the State-Trait Anxiety Inventory [STAI-5, @zsido2020development] was used as a ground truth measure of anxiety. This abridged version of the STAI [@spielberger1970manual] includes 5 items rated on a 4-point Likert scale. Changes were made in the instructions from asking *"how participants feel right now"* to *"over the past 2 weeks"* to keep it consistent with the instructions of the PHQ-4 and BDI-II. A general score of anxiety was computed by averaging all the items.

Participants were also asked to complete two questionnaires of interoception, namely the Interoceptive Accuracy Scale [IAS, 21 items rated on analog scales, @murphy2020testing] and the Multidimensional Assessment of Interoceptive Awareness [MAIA-2 - 37 items, @mehling2018multidimensional].

@@

**One of the potential limitation of our study includes the choice of the measures used as external "ground-truth" of depression and anxiety, namely the BDI-II and the STAI-5. For instance, there is ongoing debate about the STAI's  discriminant validity (which might extend to its short form used in the present study). A recent meta-analysis reported that trait anxiety scores were more strongly associated with depressive than anxiety disorders [@knowles2020specificity], suggesting that the scale may rather capture general negative affect than specifically anxiety. Regarding the BDI-II, existing evidence that suggests its relative lack of sensitivity for low-severity depression [@olino2012measuring] - comparable to similar instruments of its size (e.g., CES-D), and an improvement over the BDI-I [@wahl2014standardization] - put into question its choice for our goal of showing increased sub-clinical sensitivity. Thus, future studies should verify these findings with alternative measures of anxiety and depression and clinically assessed populations.**

Indeed, although we used a stricter criterion for classifying participants as having a depression or an anxiety disorder by restricting it to participants also reporting undergoing a medical treatment, it was still based on self-reported data. Studies in controlled clinical settings are needed to confirm the potential benefits of the refined PHQ-4 in mood disorders detection accuracy.

@@

**While adding granularity to the response format appears useful for the PHQ-4, it is possible that similar benefits could be found with other scales and measures. While the use of a limited, highly discriminating set of response options is understandable for specific applications (e.g., clinical diagnosis), we recommend future studies to investigate response format and its potential improvement for other scales used in online surveys and general population research.**
